---
title: "Working with Quebec teaching snapshots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Quebec teaching snapshots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

UlavalSSD provides fixed snapshots for practising data cleaning and exploratory
analysis. These examples run offline and use base R. The same objects can be
used with the tidyverse when those packages are installed.

```{r load}
library(UlavalSSD)
```

## Dates and missing measurements

`MeteoQuebec` retains the raw column names and types expected by the course.
Month and day are character strings. The first column, `...1`, is an export
identifier, not a measurement. Explicitly converting to a base data frame makes
subsetting behaviour independent of whether tibble is installed.

```{r weather}
weather <- as.data.frame(MeteoQuebec)
weather$date <- as.Date(with(weather, paste(year, month, day, sep = "-")))
range(weather$date)
head(weather[c("date", "min_temp", "max_temp")])
```

A complete sequence of dates does not mean every measurement is observed.
Count the missing values before choosing a summary or a model.

```{r missing}
measurements <- c(
  "max_temp", "mean_temp", "min_temp", "total_precip",
  "total_rain", "total_snow", "snow_grnd"
)
data.frame(
  variable = measurements,
  missing = vapply(weather[measurements], function(x) sum(is.na(x)), integer(1)),
  missing_proportion = vapply(
    weather[measurements], function(x) mean(is.na(x)),
    numeric(1)
  ),
  row.names = NULL
)
```

For a descriptive plot, retain dates alongside observations. Missing values
remain visible as gaps rather than being replaced by zero.

```{r plot, fig.width=6, fig.height=3.5, fig.alt="Daily minimum temperatures during 2024 in the historical teaching snapshot."}
one_year <- weather[weather$year == 2024, , drop = FALSE]
plot(one_year$date, one_year$min_temp,
  type = "l",
  xlab = "Date", ylab = "Daily minimum temperature (degrees C)"
)
```

The original weather preparation script and station-switch rule have not been
recovered. These data support cleaning exercises; a climate-trend study would
require checking station history and measurement consistency first.

## Administrative records and formatted amounts

`listecondamnation` contains historical records. It is not a representative
sample of food establishments. One establishment may appear more than once.
Counts of these records cannot estimate the probability of an offence or an
establishment's present-day operating conditions.

```{r records}
records <- as.data.frame(listecondamnation)
range(as.Date(records$Date_publication))
head(sort(table(records$Type_etablissement), decreasing = TRUE))
```

`Amende` is text. Examine its formats before parsing currency. The following
example validates a deliberately small set of known formats, rather than
silently discarding every non-numeric character in an arbitrary string.

```{r amounts}
amount_text <- c("1 500 $", "250,50 $", NA_character_)
amount_clean <- gsub("[ $]", "", amount_text)
amount_clean <- sub(",", ".", amount_clean, fixed = TRUE)
valid <- is.na(amount_clean) | grepl("^[0-9]+([.][0-9]{1,2})?$", amount_clean)
stopifnot(all(valid))
amount <- as.numeric(amount_clean)
data.frame(amount_text, amount)
```

This example uses synthetic amounts. Applying a conversion to the full dataset
requires checking all observed formats and preserving the original text for
verification.

## Exercise prompts and feedback

French output remains the default. The row feedback is a fixed answer key for
the original penguin exercise file distributed on the course website. It is
valid only before rows are filtered or reordered.

```{r prompts}
consulter_taches("histogramme", lang = "en")
verifier_valeur_aberrante(11, lang = "en")
```

## Static style feedback

Install the optional package `lintr` to use `eval_tidyverse_style()`. It examines
code without executing it. In a Quarto document it inspects fenced R chunks and
reports line numbers in the original file.

```{r style}
if (requireNamespace("lintr", quietly = TRUE)) {
  path <- tempfile(fileext = ".R")
  writeLines(c("daily_mean <- mean(c(2, 4, 6))", "print(daily_mean)"), path)
  feedback <- eval_tidyverse_style(path)
  feedback[c("total", "status", "diagnostics")]
  unlink(path)
}
```

The score is a transparent formative indicator based on seven observable
criteria. It is not a validated assessment of programming competence. Empty or
invalid code has no score, and the clarity of reasoning and usefulness of
comments remain for human review. See `?eval_tidyverse_style` for the criteria
and Quarto fence support.

## Sources and snapshot limits

Weather observations are attributed to Environment and Climate Change Canada;
administrative records are attributed to MAPAQ via Donnees Quebec. See
`?MeteoQuebec`, `?listecondamnation`, and the installed attribution file:

```{r attribution}
system.file("COPYRIGHTS", package = "UlavalSSD")
```

The datasets do not refresh during installation, loading, examples or tests.
